What is an Ancestral Recombination Graph?

### 2022-02-11 paper.tex

\end{tikzpicture}
\\
(A) & (B) \\
\end{tabular}
\caption{\label{fig-arg-data-structure}
Encoding an ARG. (A) The classical approach described by Griffiths
equates graph nodes with events in the stochastic process,
and associates recombination breakpoints with nodes. This encoding
is limited in the ancestral patterns that can be described and ambiguous.
(B) Equating graph nodes with \emph{genomes} and attaching the
interval of genome inherited by the child node with edges removes
this ambiguity and can describe any pattern of genetic inheritance.
\textcolor{red}{This is very much a draft; let's see how it works with
 the text and other figures and tidy it up in a later revision}
}
\end{figure}

@@

% it's not reasonable to assume that one type of event happens
% at a time any more.
The Griffiths ARG encoding is also limited in the types of ancestral
history that can be described. This ARG encoding is derived from the coalescent
with recombination, which explicitly assumes that only one type
of event can occur (i.e., common ancestry or recombination) at a
time. This model is derived as an approximation to a Wright-Fisher
model, assuming that the sample size $n$ is much less than the
population size $N$ and [some assumptions about genome length]
\textcolor{red}{what is this Jere?}. While these approximations are
reasonable
to make when studying the Kreitman \emph{Drosphila Melanogaster}
dataset~\citep{kreitman1983nucleotide} ($n=11$, $N\approx 10^6$,
at 43 variant sites over x Kb \textcolor{red}{details and citation
for Ne estimate}) that is classically used for benchmarking
ARG inference methods [cites] , they are highly questionable in
modern datasets. For example, there are now many
human datasets consisting of hundreds of thousands
of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and thus our sample size is often much \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
While the coalescent model is suprisingly robust to such
basic violations to the underlying assumptions~\citep{wakeley2012gene},
the combination of a large sample size and long sequence length
can produce pathological behaviour in
simulations~\citep{bhaskar2014distortion,nelson2020accounting}.

[Not sure how to make a narrative out of this, but here are the
basic problems:]
\begin{enumerate}
\item The encoding is hamstrung by the interpretation of
nodes as \emph{events} of a particular type. We can either
have common ancestor or recombination events. This is true
in the coalescent, but is not in a more general Wright-Fisher
model.
%% Note - I'm not sure this is true, if the nodes are *events* rather than
%% genomes, unless you want events (rather than genomes) to occur at
%% integer generations.
\item We can also only have one crossover per recombination event.
This is unrealistic because things like gene conversion are common. It also
means we can't have back and forth recombination between
two parents, or represent the transmission of multiple chromosomes.
\item The passage of ancestral material along edges requires
extrinsic rules which must be shared by all implementations.
These rules can be quite restrictive when generating, and
in practise some have found it simpler to add explicit
annotations to the edges.
\item Generating local trees is inefficient, essentially requiring
$O(n T)$ time (or more?)
\end{enumerate}

[WIP jk]

This ambiguity can be dealt with in a number of ways. In the original
treatments, the direction in which ancestral material travels at
a breakpoint is handled by a simple rule, sites to the left of the
breakpoint go to the left parent and sites to right go to the
right parent. [Illustrate the problem with simulating an ARG
like this]. ~\cite{ignatieva2021kwarg} resolve this problem by
associating a label ``P'' (prefix) or ``S'' (suffix) with the
outbound edges of a recombination event. This is the same approach
as used in beagle~\citep{lyngso2005minimum}.

>>>>

\end{tikzpicture}
\\
(A) & (B) \\
\end{tabular}
\caption{\label{fig-arg-data-structure}
Encoding an ARG. (A) The classical approach described by Griffiths
equates graph nodes with events in the stochastic process,
and associates recombination breakpoints with nodes. This encoding
is limited in the ancestral patterns that can be described and ambiguous.
(B) Equating graph nodes with \emph{genomes} and attaching the
interval of genome inherited by the child node with edges removes
this ambiguity and can describe any pattern of genetic inheritance.
The most important change between the two encodings is to represent
a recombination event (node 3 in A) by two parent genomes
(nodes 3 and 4 in B) and explicitly storing the interval
that the child inherits from each parent along the two corresponding
edges. Thus, child $1$ in inherits $(0, x]$ from $3$
and $(x, L)$ from $4$
(for recombination breakpoint $x$ and a sequence of length $L$).
\textcolor{red}{This is very much a draft; let's see how it works with
 the text and other figures and tidy it up in a later revision}
}
\end{figure}

@@

% it's not reasonable to assume that one type of event happens
% at a time any more.
The Griffiths ARG encoding is also limited in the types of ancestral
history that can be described. This ARG encoding is derived from the coalescent
with recombination, which explicitly assumes that only one type
of event can occur (i.e., common ancestry or recombination) at a
time. This model is derived as an approximation to a Wright-Fisher
model, assuming that the sample size $n$ is much less than the
population size $N$ and [some assumptions about genome length]
\textcolor{red}{what is this Jere?}. While these approximations are
reasonable
to make when studying the Kreitman \emph{Drosphila Melanogaster}
dataset~\citep{kreitman1983nucleotide} ($n=11$, $N\approx 10^6$,
at 43 variant sites over x Kb \textcolor{red}{details and citation
for Ne estimate}) that is classically used for benchmarking
ARG inference methods [cites],
they are highly questionable in modern datasets. % absurd really
For example, there are now many
human datasets consisting of hundreds of thousands of
genomes~\citep{bycroft2018genome,karczewski2020mutational,tanjo2021practical},
and thus our sample size is often much \emph{larger} than $N_e$
(often assumed to be $10^4$ in humans).
% TODO make this better. Ask Gregor for some refs?
% Basically, Ag datasets make the assumption of isolated
% rare recombinations absurd.
Agricultural datasets such as the 1000 bull genomes~\citep{hayes20191000}
combine genome sequence data with detailed pedigrees, with
sample data from multiple generations.
Complete chromosome-level assemblies are now possible
in humans~\citep{miga2020telomere},
and projects are under way to obtain high-quality assemblies
for all eukaryotic species in Britain and Ireland~\citep{darwin2022sequence}
and ultimately worldwide~\citep{lewin2022earth}.
While the coalescent model can be suprisingly robust to such
basic violations to the underlying assumptions~\citep{
wakeley2012gene,bhaskar2014distortion,nelson2020accounting},
it is not acceptable that our ability to \emph{represent}
ancestral relationships should be limited by these assumptions.
In a Wright-Fisher model, coalescence and recombination can
occur at the same time in the same genomes, and we can be
certain that this is also true in the incredibly rich
datasets that are available today.

Another limitation of the Griffiths ARG encoding illustated in
Fig.~\ref{fig-arg-data-structure}A is the inability to represent
multiple recombinations along a chromosome, gene conversion,
or transmission of multiple chromosomes.
[Sentence saying ``gene conversion is important, actually, cite, cite,
so we should be hoping to infer it some day''. Presumably people
already are in bacterial contexts?]
To allow us to encode data from these important processes we
must either devise workarounds to the event based encoding
(i.e., model gene conversion as two recombination events) or
generalise the model to more appropriately reflect the realities
of modern datasets.

Fortunately, it is straightforward to generalise the classical Griffiths
encoding to both simplify implementations and to encompass any
possible pattern of genetic inheritance.
We require only two changes. Firstly, we let nodes in the
graph model \emph{genomes} rather than \emph{events}. As we
see in the next section, this allows us to model complex
patterns of inheritance without classifying different types of node.
The second change we must make is to encode
intervals of ancestry inheritance \emph{explicitly} on graph edges,
rather than \emph{implicitly} as breakpoints associated with
recombination events and edge ordering rules.
That is, given two genomes $p$
and $c$ where $p$ is the parent genome and $c$ is the child
(although many generations may have elapsed between the two),
we associate a set $I$ of intervals of genome which $c$ has
inherited from $p$.
These simple changes required to translate from the
classical Griffiths approach to the proposed encoding
are shown in Fig.\ref{fig-arg-data-structure}.

\textcolor{red}{
This falls a bit flat here,  but it's not clear how we're dividing
the responsibility for explaining the model between the
different sections.
Basically, we can't explain everything at
the end of this section, but we want to give the reader the
basic idea (which is genuinely simple). Can see how things are
shaping up after we've got a draft of the rest of the
sections, I guess.}
